Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-07 tex/draft.tex
Final edit before submission

\section{Naturalness and the Higgs mass}
    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry~\cite{giudice_naturally, thooft_naturalness}. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that also explains cases where dimensional analysis fails.

@@

    There are three main approaches to dealing with the apparently unnatural Higgs mass: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass~\cite{craig}. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below~\cite{dine_naturalness}.

@@

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the uniformly null results in collider searches imply SUSY must be broken. With broken SUSY, the sum of diagrams in Fig.~\ref{exact_susy_correction} produces a correction of the form~\cite{feng}

    \fxnote{inconsistent with eq 4!!!!}  
    \noindent \begin{equation}
        \delta \mHiggs^2 = \frac{3y_t^2}{8\pi^2}\left(m_{\sfermion}^2 - m_{\f}^2 \right)\ln(\Lambda^2/m_{\sfermion}^2).
    \label{higgs_susy_correction}
    \end{equation}

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenomena. A handful of models with particular approaches to SUSY breaking that are particularly prominent or particularly relevant to naturalness or long-lived particles are gauge-mediated SUSY breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating (RPV) SUSY\fxnote{find a way to water this down or kill it}.

\section{Long-Lived Particles}
    The SUSY models most relevant to this discussion are those that are motivated by naturalness and produce long-lived particles (LLPs), a signal that could be missed by common SUSY search techniques. LLPs are defined as particles whose lifetimes are such that they decay a measurable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs~\cite{fate}. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which suppresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

    LLPs can also arise from particular SUSY mass spectra. Models in the Split SUSY paradigm, for example, propose that the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark-antiquark pair and neutralino is mediated by a highly virtual squark~\cite{kilian_split}. On the other hand, some AMSB models predict the winos as the lightest gauginos, which leads to nearly degenerate LSP and NLSP. The NLSP wino then becomes long-lived when its decay is suppressed by the lack of available phase space~\cite{randall_sundrum_out_of_this_world}.

@@

    \subsection{Tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker uses small silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves an occupancy of $\approx \SI{2}{\percent}$ with larger silicon strips~\cite{cms_tdr}.

@@

        Displaced vertex and disappearing track analyses both look for particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles within the inner volume of the detector, leaving tracks that point back to a vertex some measurable distance from the nominal interaction point (IP)~\cite{atlas_displaced}. Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed~\cite{atlas_disappearing}. Together, these two types of analyses cover a wide range of lifetimes and topologies for LLPs that decay on detector timescales.

@@

        The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector (see Fig.~\ref{material_map}) and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligible. Finally, the background from crossing tracks is determined from a model that adds fake tracks\fxnote{check} to data vertices.

@@

    \subsection{Stopped Particles}
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looks for such a signal in dimuon final states using \SI{39}{\femto\barn^{-1}} of \SI{13}{\tera\electronvolt} data~\cite{cms_stopped}.

        This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Sticking with the natural SUSY theme, this discussion will focus on long-lived gluinos, which appear in models such as Split SUSY when squarks and sleptons are quite heavy, but the gluinos are relatively light. After being pair produced, gluinos form R-hadrons and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. This process is exactly the same as in Fig.~\ref{displaced_process}, but in this case the gluino has enough time to form R-hadrons\fxnote{double check}. The gluino decay is suppressed by the large squark mass, which leads to its appreciable lifetime.

@@

        After estimating the background by extrapolating data from the cosmic-enriched region to the signal region, the authors calculate the expected background for LLP lifetimes ranging from $10^{-7}$ to $10^6$ s. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. As shown in Fig.~\ref{stopped_limits}, these results are then interpreted as upper limits on the signal production cross section and mass\fxnote{mass of what?}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    \subsection{Heavy Stable Charged Particles}
        Particles that live for more than about \SI{25}{\nano\s} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be categorized as a heavy stable charged particle (HSCP). HSCPs are characterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}~\cite{cms_hscp}.

@@

        \fxnote{I show separate trk, trk+tof limits but never explain that in text} 
        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Production cross section limits on several particles as a function of mass for the tracker-only and tracker+TOF analyses~\cite{cms_hscp}.}
        \label{hscp_limits}
        \end{center} \end{figure}

@@

    \fxnote{inconsistent w/ eq. 2?} 
    \noindent \begin{equation}
        \delta \mHiggs^2 = \mu^2
        \label{higgsino_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{3}{8\pi^2} y_t^2 m_{\tilde{t}}^2 \ln{\left(\frac{\Lambda}{Q}\right)}
        \label{stop_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{g_3^2 y_t^2}{4\pi^4} M_3^2 \ln{\left(\frac{\Lambda}{Q}\right)}^2
        \label{gluino_correction}
    \end{equation}

@@

    The uniqueness of LLP signatures and idiosyncrasy of LLP analyses leads to a broad and diverse set of analyses that will surely grow as viable natural parameter space shrinks and more imaginations are turned to filling in the gaps. Fully covering natural SUSY parameter space will take a dedicated effort from the LLP community, both to maximize the reach of individual searches and to design complementary analyses that can be used to draw general conclusions. Although natural SUSY is under pressure from conventional and long-lived null results, it remains one of the least-tuned options for natural BSM physics\fxnote{double check, then cite cornering, feng, nate, and/or giudice}. The next several years will be decisive~\cite{cornering}, and physicists will have to be clever to probe every corner of natural parameter space.

>>>>

\section{Naturalness and the Higgs mass}
    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) Any dimensionless parameter much smaller than unity must be protected by a custodial symmetry~\cite{giudice_naturally, thooft_naturalness}. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that also explains cases where dimensional analysis fails.

@@

    There are three main approaches to dealing with the apparently unnatural Higgs mass: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass~\cite{craig}. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, theories such as large extra dimensions lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck scale. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below~\cite{dine_naturalness}.

@@

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the uniformly null results in collider searches imply SUSY must be broken. With broken SUSY, the sum of diagrams in Fig.~\ref{exact_susy_correction} produces a correction of the form~\cite{craig}

    \noindent \begin{equation}
        \delta \mHiggs^2 \propto m_{\tilde{t}}^2 \ln(\Lambda/m_{\tilde{t}}).
    \label{higgs_susy_correction}
    \end{equation}

    \noindent The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenomena. Specific models will be introduced as needed throughout this discussion.

\section{Long-Lived Particles}
    The SUSY models most relevant to this discussion are those that are motivated by naturalness and produce long-lived particles (LLPs), a signal that could be missed by common SUSY search techniques. LLPs are defined as particles whose lifetimes are such that they decay a measurable distance from the nominal interaction point (IP). This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs~\cite{fate}. A similar situation arises in gauge-mediated SUSY breaking (GMSB) models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which suppresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

    LLPs can also arise from particular SUSY mass spectra. Models in the Split SUSY paradigm, for example, propose that the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark-antiquark pair and neutralino is mediated by a highly virtual squark~\cite{kilian_split}. On the other hand, some anomaly-mediated supersymmetry breaking (AMSB) models predict the winos as the lightest gauginos, which leads to nearly degenerate LSP and NLSP. The NLSP wino then becomes long-lived when its decay is suppressed by the lack of available phase space~\cite{randall_sundrum_out_of_this_world}.

@@

    \subsection{Tracker}
        In the region closest to the IP, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker uses small silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves an occupancy of $\approx \SI{2}{\percent}$ with larger silicon strips~\cite{cms_tdr}.

@@

        Displaced vertex and disappearing track analyses both look for particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles within the inner volume of the detector, leaving tracks that point back to a vertex some measurable distance from the IP~\cite{atlas_displaced}. Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed~\cite{atlas_disappearing}. Together, these two types of analyses cover a wide range of lifetimes and topologies for LLPs that decay on detector timescales.

@@

        The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector (see Fig.~\ref{material_map}) and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligible. Finally, the background from crossing tracks is determined from a model that adds fake tracksto data vertices.

@@

    \subsection{Stopped Particles}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looks for such a signal in dimuon final states using \SI{39}{\femto\barn^{-1}} of \SI{13}{\tera\electronvolt} data~\cite{cms_stopped}.

        This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Sticking with the natural SUSY theme, this discussion will focus on long-lived gluinos, which appear in models such as Split SUSY when squarks and sleptons are quite heavy, but the gluinos are relatively light. After being pair produced, gluinos form R-hadrons and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. This process is exactly the same as in Fig.~\ref{displaced_process}, but in this case the gluino has enough time to form R-hadrons. The gluino decay is suppressed by the large squark mass, which leads to its appreciable lifetime.

@@

        After estimating the background by extrapolating data from the cosmic-enriched region to the signal region, the authors calculate the expected background for LLP lifetimes ranging from $10^{-7}$ to $10^6$ s. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. As shown in Fig.~\ref{stopped_limits}, these results are then interpreted as upper limits on the signal production cross section and gluino mass, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    \subsection{Heavy Stable Charged Particles}
        Particles that live for more than about \SI{25}{\nano\s} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be categorized as a heavy stable charged particle (HSCP). HSCPs are characterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}~\cite{cms_hscp}.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_only}
        \end{center} \end{subfigure}
        \begin{subfigure}[htbp]{0.45\textwidth} \begin{center}
        \includegraphics[width=\textwidth]{images/hscp_limit_tracker_tof}
        \end{center} \end{subfigure}
        \caption{Production cross section limits on several particles as a function of mass for the tracker-only and tracker+TOF analyses~\cite{cms_hscp}.}
        \label{hscp_limits}
        \end{center} \end{figure}

@@

    \noindent \begin{equation}
        \delta \mHiggs^2 = \mu^2
        \label{higgsino_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{3}{8\pi^2} y_t^2 m_{\tilde{t}}^2 \ln{\left(\frac{\Lambda}{Q}\right)}
        \label{stop_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{g_3^2 y_t^2}{4\pi^4} M_3^2 \ln{\left(\frac{\Lambda}{Q}\right)}^2
        \label{gluino_correction}
    \end{equation}

@@

    The uniqueness of LLP signatures and idiosyncrasy of LLP analyses leads to a broad and diverse set of analyses that will surely grow as viable natural parameter space shrinks and more imaginations are turned to filling in the gaps. Fully covering natural SUSY parameter space will take a dedicated effort from the LLP community, both to maximize the reach of individual searches and to design complementary analyses that can be used to draw general conclusions. Although natural SUSY is under pressure from conventional and long-lived null results, it remains one of the least-tuned options for natural BSM physics. The next several years will be decisive~\cite{cornering}, and physicists will have to be clever to probe every corner of natural parameter space.
